---
title: "Statistical Computing"
author: "<br><br>Yanfei Kang <br> yanfeikang@buaa.edu.cn"
date: "School of Economics and Management <br> Beihang University <br> http://yanfei.site"
output:
  ioslides_presentation:
    logo: ../graphics/buaalogo.png
    widescreen: yes
subtitle: 'Lecture 3: Control structures and functions'
---

<style>
    slides > slide {
    overflow-x: auto !important;
    overflow-y: auto !important;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts = list(width.cutoff=60),
                      tidy = TRUE, 
                      echo = TRUE)
```


# Control structures

## Commonly used control structures 

- `if` and `else`: testing a condition and acting on it

- `for`: execute a loop a fixed number of times 

- `while`: execute a loop _while_ a condition is true 

- `repeat`: execute an infinite loop (must `break` out of it to stop)

- `break`: break the execution of a loop

- `next`: skip an interation of a loop

## `if`-`else`


The `if`-`else` combination is probably the most commonly used control
structure in R (or perhaps any language). For starters, you can just use the `if` statement.

```r
if(<condition>) {
        ## do something
} 
## Continue with rest of code
```

## `if`-`else`

If you have an
action you want to execute when the condition is false, then you need
an `else` clause.
```r
if(<condition>) {
        ## do something
} 
else {
        ## do something else
}
```
You can have a series of tests by following the initial `if` with any
number of `else if`s.
```r
if(<condition1>) {
        ## do something
} else if(<condition2>)  {
        ## do something different
} else {
        ## do something different
}
```

## `if`-`else` Example

```{r,prompt=FALSE}
## Generate a uniform random number
x <- runif(1, 0, 10)  
if(x > 3) {
        y <- 10
} else {
        y <- 0
}
```

Or you can write:

```{r,prompt=FALSE}
y <- if(x > 3) {
        10
} else { 
        0
}
```


## `for` Loops

For loops are most commonly used for
iterating over the elements of an object (list, vector, etc.)
```{r}
for(i in 1:10) {
        print(i)
}
```

## `for` Loops Example
The following three loops all have the same behavior.

```{r}
x <- c("a", "b", "c", "d")
for(i in 1:4) {
        ## Print out each element of 'x'
        print(x[i])  
}
```

The `seq_along()` function is commonly used in conjunction with for
loops in order to generate an integer sequence based on the length of
an object (in this case, the object `x`).

```{r}
## Generate a sequence based on length of 'x'
for(i in seq_along(x)) {   
        print(x[i])
}
```

It is not necessary to use an index-type variable. 

```{r}
for(letter in x) {
        print(letter)
}
```

## Nested `for` loops

`for` loops can be nested inside of each other.

```r
x <- matrix(1:6, 2, 3)
for(i in seq_len(nrow(x))) {
        for(j in seq_len(ncol(x))) {
                print(x[i, j])
        }   
}
```

Nested loops are commonly needed for multidimensional or hierarchical
data structures (e.g. matrices, lists). 


## `while` Loops


While loops begin by testing a condition. If it is true, then they
execute the loop body. Once the loop body is executed, the condition
is tested again, and so forth, until the condition is false, after
which the loop exits.

```{r}
count <- 0
while(count < 10) {
        print(count)
        count <- count + 1
}
```

While loops can potentially result in infinite loops if not written
properly. Use with care!




## `repeat` Loops


`repeat` initiates an infinite loop right from the start. The only way to exit a `repeat` loop is to
call `break`.


```{r,eval=FALSE,prompt=FALSE}
x0 <- 1
tol <- 1e-8
repeat {
        x1 <- computeEstimate()
        
        if(abs(x1 - x0) < tol) {  ## Close enough?
                break
        } else {
                x0 <- x1
        } 
}
```



## `next`, `break`

`next` is used to skip an iteration of a loop. 

```{r,eval=FALSE,prompt=FALSE}
for(i in 1:100) {
        if(i <= 20) {
                ## Skip the first 20 iterations
                next                 
        }
        ## Do something here
}
```

`break` is used to exit a loop immediately, regardless of what
iteration the loop may be on.

```{r,eval=FALSE,prompt=FALSE}
for(i in 1:100) {
      print(i)
      if(i > 20) {
              ## Stop loop after 20 iterations
              break  
      }		
}
```

## Summary

- Control structures like `if`, `while`, and `for` allow you to
  control the flow of an R program

- Infinite loops should generally be avoided, even if (you believe)
  they are theoretically correct.

- Control structures mentioned here are primarily useful for writing
  programs; for command-line interactive work, the "apply" functions
  are more useful.
  
# Functions

## Functions

- A transition from a mere "user" to a developer!
- Often used to
encapsulate a sequence of expressions that need to be executed
numerous times, perhaps under slightly different conditions. 
- Often written when code must be shared with others or the
public. 



## Your First Function

Functions are defined using the `function()` directive and are stored
as R objects just like anything else. In particular, they are R
objects of class "function".


```{r}
f <- function() {
        cat("Hello, world!\n")
}
f()
```

The last aspect of a basic function is the *function arguments*.

```{r}
f <- function(num) {
        for(i in seq_len(num)) {
                cat("Hello, world!\n")
        }
}
f(3)
```

## When to Write a Function?

If you find yourself doing a lot of cutting and pasting, that's usually a good sign that you might need to write a function.

This next function returns the total number of characters printed to the console.

```{r}
f <- function(num) {
        hello <- "Hello, world!\n"
        for(i in seq_len(num)) {
                cat(hello)
        }
        chars <- nchar(hello) * num
        chars
}
meaningoflife <- f(3)
print(meaningoflife)
```

## Default Values

Try this:

```{r, eval=FALSE}
f()
```

We can modify this behavior by setting a *default value* for the argument `num`. 

```{r}
f <- function(num = 1) {
        hello <- "Hello, world!\n"
        for(i in seq_len(num)) {
                cat(hello)
        }
        chars <- nchar(hello) * num
        chars
}
f()    ## Use default value for 'num'
f(2)   ## Use user-specified value
```



At this point, we have written a function that 

* has one *formal argument* named `num` with a *default value* of 1. The _formal arguments_ are the arguments included in the function definition. The `formals()` function returns a list of all the formal arguments of a function 

* prints the message "Hello, world!" to the console a number of times indicated by the argument `num`

* *returns* the number of characters printed to the console



## Argument Matching

R assigns the first value to the first argument, the second value to second argument, etc. So in the following call to `rnorm()`

```{r}
str(rnorm)
mydata <- rnorm(100, 2, 1)              ## Generate some data
```

100 is assigned to the `n` argument, 2 is assigned to the `mean` argument, and 1 is assigned to the `sd` argument, all by positional matching.


```{r}
## Positional match first argument, default for 'na.rm'
sd(mydata)                     
## Specify 'x' argument by name, default for 'na.rm'
sd(x = mydata)                 
## Specify both arguments by name
sd(x = mydata, na.rm = FALSE)  
```

When specifying the function arguments by name, it doesn't matter in what order you specify them. 

```{r}
## Specify both arguments by name
sd(na.rm = FALSE, x = mydata)     
```

You can mix positional matching with matching by name. 

```{r}
sd(na.rm = FALSE, mydata)
```

Here, the `mydata` object is assigned to the `x` argument, because it's the only argument not yet specified.

Below is the argument list for the `lm()` function, which fits linear models to a dataset.

```{r}
args(lm)
```

The following two calls are equivalent.

```r
lm(data = mydata, y ~ x, model = FALSE, 1:100)
lm(y ~ x, mydata, 1:100, model = FALSE)
```



Function arguments can also be _partially_ matched, which is useful for interactive work. The order of operations when given an argument is

1. Check for exact match for a named argument
2. Check for a partial match
3. Check for a positional match



## The `...` Argument

- There is a special argument in R known as the `...` argument, which indicate a variable number of arguments that are usually passed on to other functions. 
- The `...` argument is often used when extending another function and you don’t want to copy the entire argument list of the original function

- For example, a custom mean function may want to make use of the default `mean()` function along with its entire argument list. The function below changes the default for the `na.rm` argument to the value `na.rm = "TRUE"` (the original default was `na.rm = "FALSE"`).

```r
mymean <- function(x, na.rm = TRUE, ...) {
        mean(x, na.rm = na.rm, ...)         ## Pass '...' to 'mean?rnorm' function
}
x <- c(1, 2, NA)
mean(x)
mymean(x)
```




## Summary

* Functions can be defined using the `function()` directive and are assigned to R objects just like any other R object

* Functions have can be defined with named arguments; these function arguments can have default values

* Functions arguments can be specified by name or by position in the argument list

* Functions always return the last expression evaluated in the function body

* A variable number of arguments can be specified using the special `...` argument in a function definition.

# Looping Functions

## Looping on the Command Line

Writing `for` and `while` loops is useful when programming but not
particularly easy when working interactively on the command
line. Multi-line expressions with curly braces are just not that easy to sort through when working on the command line. R has some functions which implement looping in a compact form to make your life
easier.

- `lapply()`: Loop over a list and evaluate a function on each element 

- `sapply()`: Same as `lapply` but try to simplify the result

- `apply()`: Apply a function over the margins of an array

- `tapply()`: Apply a function over subsets of a vector

- `mapply()`: Multivariate version of `lapply`



## `lapply()`


The `lapply()` function does the following simple series of operations:

1. it loops over a list, iterating over each element in that list
2. it applies a *function* to each element of the list (a function that you specify)
3. and returns a list (the `l` is for "list"). 

This function takes three arguments: (1) a list `X`; (2) a function (or the name of a function) `FUN`; (3) other arguments via its `...` argument. If `X` is not a list, it will be coerced to a list using `as.list()`.


## `lapply()` Example 1

Here's an example of applying the `mean()` function to all elements of a list. If the original list has names, the the names will be preserved in the output.


```{r}
x <- list(a = 1:5, b = rnorm(10))
lapply(x, mean)
```

Notice that here we are passing the `mean()` function as an argument to the `lapply()` function. Functions in R can be used this way and can be passed back and forth as arguments just like any other object. When you pass a function to another function, you do not need to include the open and closed parentheses `()` like you do when you are *calling* a function.

## `lapply()` Example 2

```{r}
x <- list(a = 1:4, b = rnorm(10), c = rnorm(20, 1), d = rnorm(100, 5))
lapply(x, mean)
```

## `lapply()` Example 3

```{r}
x <- 1:4
lapply(x, runif)
```



Now how about other arguments?

Here, the `min = 0` and `max = 10` arguments are passed down to `runif()` every time it gets called.

```{r}
x <- 1:4
lapply(x, runif, min = 0, max = 10)
```



## `lapply()` Example 4

Here I am creating a list that contains two matrices.

```{r}
x <- list(a = matrix(1:4, 2, 2), b = matrix(1:6, 3, 2)) 
x
```

- How do you extract the first column of each matrix in the list?  
- Now you need an anonymous function for extracting the first column of each matrix.

```{r}
lapply(x, function(elt) { elt[,1] })
```

I can also define the function separately. 

```{r}
f <- function(elt) {
        elt[, 1]
}
lapply(x, f)
```



## `sapply()`

The `sapply()` function behaves similarly to `lapply()`; the only real difference is in the return value. `sapply()` will try to simplify the result of `lapply()` if possible. Essentially, `sapply()` calls `lapply()` on its input and then applies the following algorithm:

- If the result is a list where every element is length 1, then a vector is returned

- If the result is a list where every element is a vector of the same length (> 1), a matrix is returned.

- If it can’t figure things out, a list is returned

Here's the result of calling `lapply()`.

```{r}
x <- list(a = 1:4, b = rnorm(10), c = rnorm(20, 1), d = rnorm(100, 5))
lapply(x, mean)
```


Here's the result of calling `sapply()` on the same list.

```{r}
sapply(x, mean) 
```





## `split()`


The `split()` function takes a vector or other objects and splits it into groups determined by a factor or list of factors.

The arguments to `split()` are

```{r}
str(split)
```

where

- `x` is a vector (or list) or data frame
- `f` is a factor (or coerced to one) or a list of factors
- `drop` indicates whether empty factors levels should be dropped

The combination of `split()` and a function like `lapply()` or `sapply()` is a common paradigm in R.

## `split()` Example

Here we simulate some data and split it according to a factor variable. Note that we use the `gl()` function to "generate levels" in a factor variable.

```{r}
x <- c(rnorm(10), runif(10), rnorm(10, 1))
f <- gl(3, 10)
split(x, f)
```

A common idiom is `split` followed by an `lapply`.

```{r}
lapply(split(x, f), mean)
```

## Splitting a Data Frame

```{r}
library(datasets)
head(airquality)
```


We can split the `airquality` data frame by the `Month` variable so that we have separate sub-data frames for each month.

```{r}
s <- split(airquality, airquality$Month)
str(s)
```

Then we can take the column means for `Ozone`, `Solar.R`, and `Wind` for each sub-data frame.

```{r}
lapply(s, function(x) {
        colMeans(x[, c("Ozone", "Solar.R", "Wind")], na.rm = TRUE)
})
```

Using `sapply()` might be better here for a more readable output.

```{r}
sapply(s, function(x) {
        colMeans(x[, c("Ozone", "Solar.R", "Wind")], na.rm = TRUE)
})
```



## `tapply()`

`tapply()` is used to apply a function over subsets of a vector. It can be thought of as a combination of `split()` and `sapply()` for vectors only. 

```{r}
str(tapply)
```

The arguments to `tapply()` are as follows:

- `X` is a vector
- `INDEX` is a factor or a list of factors (or else they are coerced to factors) 
- `FUN` is a function to be applied
- ... contains other arguments to be passed `FUN`
- `simplify`, should we simplify the result?

## `tapply()` Example
Given a vector of numbers, one simple operation is to take group means.

```{r}
## Simulate some data
x <- c(rnorm(10), runif(10), rnorm(10, 1))
## Define some groups with a factor variable
f <- gl(3, 10)   
f
tapply(x, f, mean)
```

It is equivalent to:

```{r}
sapply(split(x, f), mean)
```



## `apply()`

- Used to a evaluate a function over the margins of an array. 
- Often used to apply a function to the rows or columns of a matrix. 
- Using `apply()` is not really faster than writing a loop, but it works in one line and is highly compact.


```{r}
str(apply)
```

The arguments to `apply()` are

- `X` is an array
- `MARGIN` is an integer vector indicating which margins should be “retained”. 
- `FUN` is a function to be applied
- `...` is for other arguments to be passed to `FUN`

## `apply()` Example
Here I create a 20 by 10 matrix of Normal random numbers. I then compute the mean of each column.

```{r}
x <- matrix(rnorm(200), 20, 10)
apply(x, 2, mean)  ## Take the mean of each column
```

I can also compute the sum of each row.

```{r}
apply(x, 1, sum)   ## Take the mean of each row
```

- Note that in both calls to `apply()`, the return value was a vector of numbers.
- The `MARGIN` argument essentially indicates to `apply()` which dimension of the array you want to preserve or retain. So when taking the mean of each column, I specify

```{r,eval=FALSE}
apply(x, 2, mean)
```


## Col/Row Sums and Means

For the special case of column/row sums and column/row means of matrices, we have some useful shortcuts.

- `rowSums` = `apply(x, 1, sum)`
- `rowMeans` = `apply(x, 1, mean)`
- `colSums` = `apply(x, 2, sum)`
- `colMeans` = `apply(x, 2, mean)`

The shortcut functions are heavily optimized and hence are _much_ faster, but you probably won’t notice unless you’re using a large matrix. Another nice aspect of these functions is that they are a bit more descriptive. It's arguably more clear to write `colMeans(x)` in your code than `apply(x, 2, mean)`.

## Other Ways to Apply

You can do more than take sums and means with the `apply()` function. For example, you can compute quantiles of the rows of a matrix using the `quantile()` function.

```{r}
x <- matrix(rnorm(200), 20, 10)
## Get row quantiles
apply(x, 1, quantile, probs = c(0.25, 0.75))    
```

Notice that I had to pass the `probs = c(0.25, 0.75)` argument to `quantile()` via the `...` argument to `apply()`.




## `mapply()`

The `mapply()` function is a multivariate apply of sorts which applies a function in parallel over a set of arguments. Recall that `lapply()` and friends only iterate over a single R object. What if you want to iterate over multiple R objects in parallel? This is what `mapply()` is for.

```{r}
str(mapply)
```

The arguments to `mapply()` are

- `FUN` is a function to apply
- `...` contains R objects to apply over
- `MoreArgs` is a list of other arguments to `FUN`.
- `SIMPLIFY` indicates whether the result should be simplified

The `mapply()` function has a different argument order from `lapply()` because the function to apply comes first rather than the object to iterate over. The R objects over which we apply the function are given in the `...` argument because we can apply over an arbitrary number of R objects.

## `mapply()` Example
For example, the following is tedious to type

`list(rep(1, 4), rep(2, 3), rep(3, 2), rep(4, 1))`

With `mapply()`, instead we can do

```{r}
 mapply(rep, 1:4, 4:1)
```

This passes the sequence `1:4` to the first argument of `rep()` and the sequence `4:1` to the second argument.


Here's another example for simulating randon Normal variables.

```{r}
noise <- function(n, mean, sd) {
      rnorm(n, mean, sd)
}
## Simulate 5 randon numbers
noise(5, 1, 2)        
```

Here we can use `mapply()` to pass the sequence `1:5` separately to the `noise()` function so that we can get 5 sets of random numbers, each with a different length and mean.


```{r}
x <- mapply(noise, 1000, 1:5, 1:5)
apply(x, 2, mean)
```

The above call to `mapply()` is the same as

```{r}
list(noise(1000, 1, 1), noise(1000, 2, 2),
     noise(1000, 3, 3), noise(1000, 4, 4),
     noise(1000, 5, 5))
```

## Vectorizing a Function

- The `mapply()` function can be use to automatically "vectorize" a function. 
- It can be used to take a function that typically only takes single arguments and create a new function that can take vector arguments. 

Here's an example of a function that computes the sum of squares $\sum_{i=1}^n(x_i-\mu)^2/\sigma^2$.

```{r}
sumsq <- function(mu, sigma, x) {
        sum(((x - mu) / sigma)^2)
}
```

This function takes a mean `mu`, a standard deviation `sigma`, and some data in a vector `x`.

In many statistical applications, we want to minimize the sum of squares to find the optimal `mu` and `sigma`. Before we do that, we may want to evaluate or plot the function for many different values of `mu` or `sigma`. However, passing a vector of `mu`s or `sigma`s won't work with this function because it's not vectorized.

```{r}
x <- rnorm(100)       ## Generate some data
sumsq(1:10, 1:10, x)  ## This is not what we want
```

Note that the call to `sumsq()` only produced one value instead of 10 values.

However, we can do what we want to do by using `mapply()`.

```{r}
mapply(sumsq, 1:10, 1:10, MoreArgs = list(x = x))
```

There's even a function in R called `Vectorize()` that automatically can create a vectorized version of your function. So we could create a `vsumsq()` function that is fully vectorized as follows.

```{r}
vsumsq <- Vectorize(sumsq, c("mu", "sigma"))
vsumsq(1:10, 1:10, x)
```




## Summary

* The loop functions in R are very powerful because they allow you to conduct a series of operations on data using a compact form

* The operation of a loop function involves iterating over an R object (e.g. a list or vector or matrix), applying a function to each element of the object, and the collating the results and returning the collated results.

* Loop functions make heavy use of anonymous functions, which exist for the life of the loop function but are not stored anywhere

* The `split()` function can be used to divide an R object in to subsets determined by another variable which can subsequently be looped over using loop functions.

# Lab Session 2

## 
In this lab, you will use the temperature data in four cities: Melbourne, Sydney, Brisbane and Cairns. You can download them from https://yanfei.site/docs/sc/data/temp.zip. 

1. Please make a function `load.file()` to read a .csv file and transform the first column (a character representing date and time) using `as.POSIXlt` into R time format.
2. Then apply `load.file()` to each filename using `lapply()`.
3. How many rows of data are there for each city?
4. What is the hottest temperature recorded by city?
5. Estimate the autocorrelation function for each city.



## References

Chapters  14, 15 and 18 of the book "R programming for data science".
