---
title: "Statistical Computing"
author: "<br><br>Yanfei Kang <br> yanfeikang@buaa.edu.cn"
date: "School of Economics and Management <br> Beihang University <br> http://yanfei.site"
output:
  ioslides_presentation:
    logo: ../graphics/buaalogo.png
    widescreen: yes
subtitle: 'Lecture 1: R basics'
---

<style>
    slides > slide {
    overflow-x: auto !important;
    overflow-y: auto !important;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts = list(width.cutoff=60),
                      tidy = TRUE, 
                      echo = TRUE,
                      collapse = TRUE)
```

## Objectives 

- Overview of R
- R nuts and bolts
- Getting data in and out of R
- Subsetting R objects

# Overview of R

## What is R?

> - A freely available language and environment 
> - Statistical computing and graphics 
> - Linear and nonlinear modelling, statistical tests, time series analysis, classification, clustering, etc.

## Installation

- [Install R](https://cran.r-project.org/)
- [Install Rstudio](https://www.rstudio.com/products/rstudio/download/#download)

<br>
**Why Rstudio?**

- Syntax highlighting
- Able to evaluate R code
      + by line
      + by selection 
      + entire file
- Command auto-completion



## Design of the R System

- When you download R from CRAN, you get the "base" system - a substantial amount of functionality.
- > 10,000 packages on CRAN that have been developed by users and programmers around the world.
- People often make packages available on their personal websites.
- There are a number of packages being developed on repositories like GitHub and BitBucket.


# R Nuts and Bolts

## Basic Operations

```{r}
1 + 2 + 3
1 + 2 * 3 

x <- 1
y <- 2
z <- c(x,y)
z

exp(1)
cos(3.141593)
log2(1)
```

## R Objects

R has five basic classes of objects:

1. character
2. numeric (real numbers)
3. integer
4. complex
5. logical (True/False)

## Numbers

- Numbers in R are generally treated as numeric objects.
- Difference of `1` and `1L`?
- Special number `Inf`. Try `1/Inf`.
- `NaN`: an undefined value (not a number). Try `0/0`. It can also be thought of as a missing value.

## Attributes

Attributes can be accessed by `attributes()`. Some examples of R object attributes are:

- names, dimnames
- dimensions (e.g. matrices, arrays)
- class (e.g. integer, numeric)
- length


## Vectors

The `c()` function can be used to create vectors of objects by concatenating things together. 

```{r}
x <- c(0.5, 0.6)       ## numeric
x <- c(TRUE, FALSE)    ## logical
x <- c(T, F)           ## logical
x <- c("a", "b", "c")  ## character
x <- 9:29              ## integer
x <- c(1+0i, 2+4i)     ## complex
```

You can also use the  `vector()` function to initialize vectors.

```{r}
x <- vector("numeric", length = 10) 
x
```


## Matrices

```{r}
m <- matrix(c(1:6), 2, 3)
attributes(m)
dim(m)
t(m)
m[1, 2]
m[1, ]
n <- matrix(c(8:13), 2, 3)
cbind(m, n)
rbind(m, n)
```

## Lists

- Special data structure that matrix could not handle.
    - Data length are not the same.
    - Data type are not the same.

```{r}
l <- list(a = c(1, 2), b = 'apple')
attributes(l)
```

## Factors

Factors are used to represent categorical data.
```{r}
f <- factor(c("yes", "yes", "no", "yes", "no")) 
attributes(f)
```


## Data Frames

- A special type of list.
- Unlike matrices -- data frames can store different classes of objects in each column.
- They have column names and row names.

```{r}
d <- data.frame(x = 1:10, y = letters[1:10])
attributes(d)
names(d)
row.names(d)
```

## Names

Names are very useful for writing readable code and self-describing objects.
```{r}
x <- 1:3
names(x)
names(x) <- c("New York", "Seattle", "Los Angeles") 
x
names(x)
```
Lists can also have names, which is often very useful.
```{r}
x <- list("Los Angeles" = 1, "Boston" = 2, "London"= 3) 
x
names(x)
```

# Getting Data in and out of R

## Reading and Writing Data

There are a few principal functions reading data into R. 

* `read.table`, `read.csv`, for reading tabular data 
* `readLines`, for reading lines of a text file
* `source`, for reading in R code files (`inverse` of `dump`) 
* `dget`, for reading in R code files (`inverse` of `dput`)
* `load`, for reading in saved workspaces

There are analogous functions for writing data to files.

* `write.table`, for writing tabular data to text files (i.e. CSV) or
  connections
* `writeLines`, for writing character data line-by-line to a file or
  connection
* `dump`, for dumping a textual representation of multiple R objects
* `dput`, for outputting a textual representation of an R object
* `save`, for saving an arbitrary number of R objects in binary format
  (possibly compressed) to a files


There are many R packages that have been developed to read
in all kinds of other datasets (e.g., the `readr` package).

# Subsetting R objects

## How to Subset?

There are three operators that can be used to extract subsets of R
objects.

- The `[` operator always returns an object of the same class as the
  original. It can be used to select multiple elements of an object

- The `[[` operator is used to extract elements of a list or a data
  frame. It can only be used to extract a single element and the class
  of the returned object will not necessarily be a list or data frame.

- The `$` operator is used to extract elements of a list or data frame
  by literal name. Its semantics are similar to that of `[[`.
  
## Subsetting a Vector
Vectors are basic objects in R and they can be subsetted using the `[`
operator.

```{r}
x <- c("a", "b", "c", "c", "d", "a")  
x[1]    ## Extract the first element
x[2]    ## Extract the second element
```

The `[` operator can be used to extract multiple elements of a vector
by passing the operator an integer sequence. Here we extract the first
four elements of the vector.

```{r}
x[1:4]
x[c(1, 3, 4)]
x[x > 2]
```

## Subsetting a Matrix


Matrices can be subsetted in the usual way with (_i,j_) type
indices. 

```{r}
x <- matrix(1:6, 2, 3)
x
```

We can access the $(1,2)$ or the $(2,1)$ element of this matrix
using the appropriate indices.

```{r}
x[1, 2]
x[2, 1]
```

Indices can also be missing. This behavior is used to access entire
rows or columns of a matrix.

```{r}
x[1, ]  ## Extract the first row
x[, 2]  ## Extract the second column
```

## Subsetting Lists
ists in R can be subsetted using all three of the operators mentioned
above, and all three are used for different purposes.

```{r}
x <- list(foo = 1:4, bar = 0.6)
x
```

The `[[` operator can be used to extract *single* elements from a
list. Here we extract the first element of the list.

```{r}
x[[1]]
```

The `[[` operator can also use named indices so that you don't have to
remember the exact ordering of every element of the list. You can also
use the `$` operator to extract elements by name.

```{r}
x[["bar"]]
x$bar
```

## Subsetting Nested Elements of a List

The `[[` operator can take an integer sequence if you want to extract
a nested element of a list.

```{r}
x <- list(a = list(10, 12, 14), b = c(3.14, 2.81))
## Get the 3rd element of the 1st element
x[[c(1, 3)]]  
## Same as above
x[[1]][[3]]   
## 1st element of the 2nd element
x[[c(2, 1)]]  
```

## Extracting Multiple Elements of a List


The `[` operator can be used to extract *multiple* elements from a
list. For example, if you wanted to extract the first and third
elements of a list, you would do the following

```{r}
x <- list(foo = 1:4, bar = 0.6, baz = "hello")
x[c(1, 3)]
```

Note that `x[c(1, 3)]` is NOT the same as `x[[c(1, 3)]]`.

Remember that the `[` operator always returns an object of the same
class as the original. Since the original object was a list, the `[`
operator returns a list. In the above code, we returned a list with
two elements (the first and the third).

## Removing NA Values
A common task in data analysis is removing missing values (`NA`s).

```{r}
x <- c(1, 2, NA, 4, NA, 5)
bad <- is.na(x)
print(bad)
x[!bad]
```
What if there are multiple R objects and you want to take the subset with
no missing values in any of those objects?

```{r}
head(airquality)
good <- complete.cases(airquality)
head(airquality[good, ])
```

## Review of this lecture

- Overview of R
- R nuts and bolts
- Getting data in and out of R
- Subsetting R objects

# Lab Session 1

## Read and Write Data in R

You'll be working with [swimming_pools.csv](http://s3.amazonaws.com/assets.datacamp.com/production/course_1477/datasets/swimming_pools.csv); it contains data on swimming pools in Brisbane, Australia (Source: [data.gov.au](https://data.gov.au/)). The file contains the column names in the first row. It uses a comma to separate values within rows.

1. Try `read.csv()` and `read.table()` to import "swimming_pools.csv" as a data frame with the name `pools`.
2. Try `write.table()`, `dput()`, and `save()` functions to write `pools` to files.
3. Restart R and read your saved data in R.
4. Practice subsetting of a data frame.

## References

### Chapters 3-10 of the book "R programming for data science".